Semi-quantitative petrographic characterization of Bronze Age quartzite tools – a multivariate approach

### 2021-12-22 analysis/paper/paper.Rmd
added Aims section

In the current study a dataset of quartz and quartzite artefacts from museum collections in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman, XRF). The aim is to evaluate their viability for fingerprinting quartz and quartzite material in the North Swedish region, where there currently is a lack of reference material as well as a complex geological setting. This will be achieved via Principal Component Analysis (PCA), where the observed groupings of the material in the model will be related to their spectral features. 
 \linebreak

# Study region

@@

In 1925 the Geological Society arranged an exursion to Södra Storfjället in Västerbotten. In the description of the regional stratigraphy it is noted that the lower parts are made up of, among others, "*...dark schists and graphitic quartzite...*" [@geologiska25 p.147]. Apart from these quartzite with little impurities and Alum shale are also noted.

# Material
```{r select-point-data}
#Filter data to focus on points, point fragments and preforms made from quartz/quartzite materials
Points2 <-
  samples %>%
    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
         type == "Point" | type == "Point fragment" | type == "Preform")

>>>>

##Aims

In the current study a dataset of quartz and quartzite artefacts from museum collections in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman, XRF). The aim of the study is to evaluate their viability for fingerprinting quartz and quartzite material in the North Swedish region, where there currently is a lack of reference material as well as a complex geological setting. The study is a part of a larger project wherein models of settlement and mobility are being evaluated and developed in regards to raw material procurement and use. As flint is a non-local material to Northern Sweden it follows that any find of this material within the region most likely have been brought there through anthropogenic activity. In contrast, determining the degree to which the distribution pattern of quartz and quartzite have been influenced by cultural and/or environmental factors however, is far more difficult. While there is a good understanding of the geological makeup of the Scandes [@stephens_bergmanweihed20], the quartzite itself remains largely uncharacterized, with few reference samples available for comparison. The glacial impact on the landscape further compounds the issue as geological material has been transported and deposited by glacial movements, as well as subsequent postglacial processes. Although models for settlement and mobility can be used to test ideas of material distribution as influenced by cultural processes [e.g. @forsberg85], it is also necessary to attempt to estimate the degree of environmental impact on this pattern.

This will be achieved via Principal Component Analysis (PCA), where the observed groupings of the material in the model will be related to their spectral features.

<!--# Study region

@@

In 1925 the Geological Society arranged an exursion to Södra Storfjället in Västerbotten. In the description of the regional stratigraphy it is noted that the lower parts are made up of, among others, "*...dark schists and graphitic quartzite...*" [@geologiska25 p.147]. Apart from these quartzite with little impurities and Alum shale are also noted. 
-->
# Material
```{r select-point-data}
#Filter data to focus on points, point fragments and preforms made from quartz/quartzite materials
Points2 <-
  samples %>%
    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
         type == "Point" | type == "Point fragment" | type == "Preform")
